---
title: "Modeling"
author: "Sofia Perez"
date: "2026-05-21"
format: html
editor: visual
toc: true # use this to display a table of contents
execute:
  message: false # use this to make sure messages don't show up
  warning: false # use this to make sure warnings don't show up
---

## setup

```{r}
#| label: packages and data
#| message: false
# read in your packages here
library(tidyverse) # general use
library(here) # file organization
library(janitor) # cleaning data frames
library(readxl) # reading excel files
library(scales) # modifying axis labels
library(ggeffects) # getting model predictions
library(MuMIn) # model selection
library(showtext) # to show custom text font on plots
library(flextable) # create tables for model summaries

font_add_google(name = "Public Sans", # name of font on Google font website
              family = "Psans") # name to call in font as
showtext_auto()

# read in your data here
labrynth <- read_csv(here("data",
                             "LagoonLappers_DataSheetES153 - R-Studio Data.csv"))
# clean data
labrynth_clean <- labrynth |> # create new object
  clean_names() # clean column names to lower case and underscores
```

# Figures

## Hypothesis 1.

**Long-term daily direct exposure to sunlight on a patch of vegetation will show a higher flower count compared to vegetation characterized by shadows from trees.**

```{r}
#| label: linear regression shadow
shadow_lm <- lm(formula = flower_bud_count ~ shadow_from_tree_root_ft,
                data = labrynth_clean)

flextable::as_flextable(shadow_lm)
```

```{r}
#| label: correlation shadow
cor.test(labrynth_clean$flower_bud_count, labrynth_clean$shadow_from_tree_root_ft, # run a correlation test to check strength between salinity and biomass
         method = "pearson") |>  # specifcy pearson rank correlation
  print()
```

```{r}
#| label: linear model shadow
model_pred1 <- ggpredict(
  shadow_lm,
  terms = "shadow_from_tree_root_ft [14.95:17.05 by = 0.01]"
)
```

### Figure 1. Sunlight Exposure and Flower Count

```{r}
#| label: Shadow and flower plot
#| warning: false
ggplot(data = labrynth_clean,
       aes(x = shadow_from_tree_root_ft,
       y = flower_bud_count)) +
  geom_ribbon(data = model_pred1, # plot prediction 95% CI using mod_preds object
              aes(x = x, # use same x-axix
                  y = predicted, # set y-axis as predicted values
                  ymin = conf.low, # set bottom of region
                  ymax = conf.high), # set top
              alpha = 0.2, # adjust tranparency
              fill = "firebrick2") + # change color
  geom_line(data = model_pred1, # plot model prediction as line using mod_preds object
            aes(x = x, # use same x-axis
                y = predicted), # set y-axis as predicted values
            linewidth = 1, # change line width to thicker 
            color = "firebrick4") + # change color
  geom_point(size = 3.5,
             alpha = 0.5,
             aes(color = common_name_plant)) +
  
  
  labs(title = "Flower bud patterns by tree shadow",
       x = "Shadow length from tree root (ft)",
       y = "Flower bud count",
       color = "Species") +
  theme_light(base_family = "Psans") + # change theme and font
  theme(legend.position = "right",
        plot.title = element_text(hjust = 0.5))
```

## Hypothesis 2.

**Native plant vegetation will be shorter if there is taller non-native grasses in close proximity.**

```{r}
#| label: Coyote Brush linear regression
coyote <- labrynth_clean |> 
  filter(common_name_plant == "*Coyote Brush") # only include rows where `jtnp` is in the `park` column

coyote_lm <- lm(formula = height_of_each_plant_species_in ~ date,
                data = coyote)

flextable::as_flextable(coyote_lm)
```

```{r}
#| label: Western Ragweed linear regression
ragweed <- labrynth_clean |> 
  filter(common_name_plant == "*Western Ragweed") # only include rows where `Western Ragweed` is in the species column

ragweed_lm <- lm(formula = height_of_each_plant_species_in ~ date,
                data = ragweed)

flextable::as_flextable(ragweed_lm)
```

```{r}
#| label: Great Brome Grass linear regression
grass <- labrynth_clean |> 
  filter(common_name_plant == "Great Brome Grass") # only include rows where `Great Brome Grass` is in the species column

grass_lm <- lm(formula = height_of_each_plant_species_in ~ date,
                data = grass)

flextable::as_flextable(grass_lm)
```

```{r}
#| label: Scaret Pimpernel linear regression
scarlet <- labrynth_clean |> 
  filter(common_name_plant == "Scarlet Pimpernel") # only include rows where `Scarlet Pimpernel` is in the species column

scarlet_lm <- lm(formula = height_of_each_plant_species_in ~ date,
                data = scarlet)

flextable::as_flextable(scarlet_lm)
```

```{r}
#| label: Wild Radish linear regression
radish <- labrynth_clean |> 
  filter(common_name_plant == "Wild Radish") # only include rows where `Scarlet Pimpernel` is in the species column

radish_lm <- lm(formula = height_of_each_plant_species_in ~ date,
                data = radish)

flextable::as_flextable(radish_lm)
```

### Figure 2. Species height over time

```{r}
#| label: species height over time plot
#| warning: false
ggplot(data = labrynth_clean,
       aes(x = date,
           y = height_of_each_plant_species_in,
           color = common_name_plant)) +
  geom_point(size = 3.5) +
  geom_smooth(method = "lm",
              show.legend = FALSE,
              se = FALSE) +
  theme_light(base_family = "Psans") + # change theme and font
  labs(title = "Plant growth by species over time",
       x = "Date",
       y = "Plant height (in)",
       color = "Species") +
  theme(legend.position = "right",
        plot.title = element_text(hjust = 0.5))
```

## Hypothesis 3.

**Consistently strong winds will tilt and decrease overall growth height of plants in our study system.**

```{r}
#| label: linear regression wind
wind_lm <- lm(formula = height_of_each_plant_species_in ~ wind_speed_mph,
              data = labrynth_clean)

flextable::as_flextable(wind_lm)
```

```{r}
#| label: correlation wind
cor.test(labrynth_clean$height_of_each_plant_species_in, labrynth_clean$wind_speed_mph, # run a correlation test to check strength between salinity and biomass
         method = "pearson") |>  # specifcy pearson rank correlation
  print()
```

```{r}
#| label: linear model wind
model_pred2 <- ggpredict(
  wind_lm,
  terms = "wind_speed_mph [5.5:20.5 by = 0.1]"
) 
```

### Figure 3. Wind and plant height

```{r}
#| label: Wind speed and plant height
ggplot(data = labrynth_clean,
       aes(x = wind_speed_mph,
           y = height_of_each_plant_species_in)) +
  geom_ribbon(data = model_pred2, # plot prediction 95% CI using mod_preds object
              aes(x = x, # use same x-axix
                  y = predicted, # set y-axis as predicted values
                  ymin = conf.low, # set bottom of region
                  ymax = conf.high), # set top
              alpha = 0.2, # adjust tranparency
              fill = "firebrick2") + # change color
  geom_line(data = model_pred2, # plot model prediction as line using mod_preds object
            aes(x = x, # use same x-axis
                y = predicted), # set y-axis as predicted values
            linewidth = 1, # change line width to thicker 
            color = "firebrick4") +
  geom_point(size = 3.5,
             aes(color = common_name_plant)) +
  theme_light(base_family = "Psans") + # change theme and font
  labs(title = "Wind speed relationship with plant growth",
       x = "Wind speed (mph)",
       y = "Plant height (in)",
       color = "Species") +
  theme(legend.position = "right",
        plot.title = element_text(hjust = 0.5))
```

:P
