---
title: "Modeling"
author: "Sofia Perez"
date: "2026-05-21"
format: html
editor: visual
---

## setup

```{r}
#| label: packages and data
#| message: false
# read in your packages here
library(tidyverse) # general use
library(here) # file organization
library(janitor) # cleaning data frames
library(readxl) # reading excel files
library(scales) # modifying axis labels
library(ggeffects) # getting model predictions
library(MuMIn) # model selection
library(showtext) # to show custom text font on plots

font_add_google(name = "Public Sans", # name of font on Google font website
              family = "Psans") # name to call in font as
showtext_auto()

# read in your data here
labrynth <- read_csv(here("data",
                             "LagoonLappers_DataSheetES153 - R-Studio Data.csv"))
# clean data
labrynth_clean <- labrynth |> # create new object
  clean_names() # clean column names to lower case and underscores
```

# Figures

**Hypothesis 1.** Long-term daily direct exposure to sunlight on a patch of vegetation will show a higher flower count compared to vegetation characterized by shadows from trees.

```{r}
#| label: linear regression shadow
shadow_lm <- lm(formula = flower_bud_count ~ shadow_from_tree_root_ft, data = labrynth_clean)
```

```{r}
#| label: linear model shadow
model_pred1 <- ggpredict(
  shadow_lm,
  terms = "shadow_from_tree_root_ft [14.95:17.05 by = 0.01]"
) 
```

## Figure 1. Sunlight Exposure and Flower Count

```{r}
#| label: Shadow and flower plot
#| warning: false
ggplot(data = labrynth_clean,
       aes(x = shadow_from_tree_root_ft,
       y = flower_bud_count)) +
  geom_ribbon(data = model_pred1, # plot prediction 95% CI using mod_preds object
              aes(x = x, # use same x-axix
                  y = predicted, # set y-axis as predicted values
                  ymin = conf.low, # set bottom of region
                  ymax = conf.high), # set top
              alpha = 0.2, # adjust tranparency
              fill = "firebrick2") + # change color
  geom_line(data = model_pred1, # plot model prediction as line using mod_preds object
            aes(x = x, # use same x-axis
                y = predicted), # set y-axis as predicted values
            linewidth = 2, # change line width to thicker 
            color = "firebrick4") + # change color
  geom_point(size = 3.5,
             alpha = 0.5,
             aes(color = common_name_plant)) +
  
  
  labs(title = "Flower bud patterns by tree shadow",
       x = "Shadow length from tree root (ft)",
       y = "Flower bud count",
       color = "Species") +
  theme_light(base_family = "Psans") + # change theme and font
  theme(legend.position = "right",
        plot.title = element_text(hjust = 0.5))
```

**Hypothesis 2.** Native plant vegetation will be shorter if there is taller non-native grasses in close proximity.

## Figure 2. Species height over time

```{r}
#| label: species height over time plot
ggplot(data = labrynth_clean,
       aes(x = date,
           y = height_of_each_plant_species_in,
           color = common_name_plant)) +
  geom_point(size = 3.5) +
  theme_light(base_family = "Psans") + # change theme and font
  labs(title = "Plant growth by species over time",
       x = "Date",
       y = "Plant height (in)",
       color = "Species") +
  theme(legend.position = "right",
        plot.title = element_text(hjust = 0.5))
```

**Hypothesis 3.** Consistently strong winds will tilt and decrease overall growth height of plants in our study system.

## Figure 3. Wind and plant height

```{r}
#| label: Wind speed and plant height
ggplot(data = labrynth_clean,
       aes(x = wind_speed_mph,
           y = height_of_each_plant_species_in,
           color = common_name_plant)) +
  geom_point(size = 3.5) +
  theme_light(base_family = "Psans") + # change theme and font
  labs(title = "Wind speed relationship with plant growth",
       x = "Wind speed (mph)",
       y = "Plant height (in)",
       color = "Species") +
  theme(legend.position = "right",
        plot.title = element_text(hjust = 0.5))
```

**Extra figure** – yadayada
